Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-05-09 article-models-ttderived.tex
rewritten the part about the second judgmental equality of extension algebras

>>>>

\subsection{Extension algebras}
In this subsection our goal is to define the notion of extension algebras,
which are internal versions of the extension operation of the theory of
contexts, families and terms. In this article, their use will be mainly in
universes.

Let $P$ be a family over an extended context $\ctxext{\Gamma}{A}$. We could
mimic extension by requiring to have terms
\begin{align*}
\jhom*{\Gamma}{{A}{P}}{A}{e}\\
\jhom*{{\Gamma}{A}}{{P}{\jcomp{}{e}{P}}}{P}{f}.
\end{align*}
Extension also satisfies the properties explained in \autoref{comp-ee}, so we
must find the two judgmental equalities for $e$ and $f$ mimicing those. 
The first of these judgmental equalities is easy to give: it says that the
following diagram commutes:
\begin{equation}\label{eq:ealg-eq1}
\begin{tikzcd}[column sep=huge]
\ctxext{A}{{P}{\jcomp{}{e}{P}}} 
  \ar{d}[swap]{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}
    %\tmext{\ctxwk{\jcomp{}{e}{P}}{e}}{\idtm{\jcomp{}{e}{P}}}
    } 
  \ar{r}{\jvcomp{}{\idtm{A}}{f}
    %\tmext{\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}{f}
    } 
  & \ctxext{A}{P} \ar{d}{e}\\
\ctxext{A}{P} \ar{r}[swap]{e} & A
\end{tikzcd}
\end{equation}
To get a feel for this judgmental equality we include the following lemma.

\begin{lem}
Let $A$, $P$, $e$ and $f$ be as above, satisfying \autoref{eq:ealg-eq1} and
let $x_0:A$,
$x_1:\subst{x_0}{P}$ and $x_2:\subst{x_1}{{x_0}{\jcomp{}{e}{P}}}$.
Then we have the judgmental equality
\begin{equation*}
\subst{{x_2}{{x_1}{f}}}{{x_0}{e}}
\jdeq
\subst{x_2}{{{x_1}{{x_0}{e}}}{e}}.
\end{equation*}
\end{lem}

\begin{proof}
The proof is a simple computation:
\begin{align*}
\subst{{x_2}{{x_1}{f}}}{{x_0}{e}}
& \jdeq
  \subst{\tmext{x_0}{\subst{x_2}{{x_1}{f}}}}{e}
  \\
& \jdeq 
  \subst{{x_2}{{x_1}{{x_0}{\jvcomp{}{\idtm{A}}{f}}}}}{e}
  \\
& \jdeq
  \subst{x_2}{{x_1}{{x_0}{\jcomp{}{\jvcomp{}{\idtm{A}}{f}}{e}}}}
  \\
& \jdeq
  \subst{x_2}{{x_1}{{x_0}{\jcomp{}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}{e}}}}
  \\
& \jdeq 
  \subst{{x_2}{{x_1}{{x_0}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}}}}{e}
  \\
& \jdeq
  \subst{\tmext{\subst{x_1}{{x_0}{e}}}{x_2}}{e}
  \\
& \jdeq
  \subst{x_2}{{{x_1}{{x_0}{e}}}{e}}.\qedhere
\end{align*}
\end{proof}

The second of the judgmental equalities is harder to describe, however. We need
to consider `higher' families, i.e.~families over families over families, and
thus we need to look at the family $\jcomp{}{e}{\jcomp{}{e}{P}}$ and find the
two dotted morphisms in the diagram
\begin{equation*}
\begin{tikzcd}
\ctxext{P}{{\jcomp{}{e}{P}}{\jcomp{}{e}{\jcomp{}{e}{P}}}}
  \ar[densely dotted]{d}
  \ar[densely dotted]{r}
& \ctxext{P}{\jcomp{}{e}{P}} \ar{d}{f}\\
\ctxext{P}{\jcomp{}{e}{P}} \ar{r}{f} & P
\end{tikzcd}
\end{equation*}
The first is easy to find. Note that we have the judgmental equality
\begin{equation*}
\ctxext{\jcomp{}{e}{P}}{\jcomp{}{e}{\jcomp{}{e}{P}}}
  \jdeq
  \jcomp{}{e}{\ctxext{P}{\jcomp{}{e}{P}}}
\end{equation*}
and therefore we may just take the morphism
\begin{equation*}
\jhom
  {{\Gamma}{A}}
  {{P}{{\jcomp{}{e}{P}}{\jcomp{}{e}{\jcomp{}{e}{P}}}}}
  {{P}{\jcomp{}{e}{P}}}
  {\jvcomp{}{\idtm{P}}{\jcomp{}{e}{f}}}.
\end{equation*}
For the other morphism we need to look at the family
$\ctxext{P}{{\jcomp{}{e}{P}}{\jcomp{}{e}{\jcomp{}{e}{P}}}}$ differently. We do
that in the following lemma.

\begin{lem}
Suppose we have $A$, $P$, $e$ and $f$ satisfying \autoref{eq:ealg-eq1}. Then
we have the judgmental equality
\begin{equation*}
\jcomp{}{e}{\jcomp{}{e}{P}}
  \jdeq
  \jcomp{}{f}{\jcomp{}{e}{P}}.
\end{equation*}
\end{lem}

\begin{proof}
The proof is a rather long computation:
\begin{align*}
\jcomp{}{e}{\jcomp{}{e}{P}}
& \jdeq
  \subst{e}
    {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { \ctxwk
        {\jcomp{}{e}{P}}{\subst{e}
        {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\jcomp{}{e}{P}}{{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\ctxext{{A}{P}}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \subst
    {\unfold{\jvcomp{\jcomp{}{e}{P}}{e}{\idtm{\jcomp{}{e}{P}}}}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}
  \\
& \jdeq 
  \jcomp{}{\jcomp{}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}{e}}{P}
  \\
& \jdeq
  \jcomp{}{
    \jcomp{}{\jvcomp{}{\idtm{A}}{f}}{e}}{P}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{\idtm{A}}{f}}{
    \jcomp{}{e}{P}}
  \\
& \jdeq
  \subst
    {\tmext{\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}{f}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {f}
    { {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
      }
  \\
& \jdeq
  \subst
    {f}
    { {\cprojfstf{A}{\ctxext{P}{\jcomp{}{e}{P}}}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
      }
  \\
& \jdeq
  \subst
    {f}
    {\jcomp{}{\cprojfstf{A}{\ctxext{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {f}
    {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{f}{\jcomp{}{e}{P}}.
  \qedhere
\end{align*}
\end{proof}

Now we see that we can use the morphism
\begin{equation*}
\jhom
  {{\Gamma}{A}}
  {{P}{{\jcomp{}{e}{P}}{\jcomp{}{f}{{}{e}{P}}}}}
  {{P}{\jcomp{}{e}{P}}}
  {\jvcomp{}{f}{\idtm{\jcomp{}{f}{{}{e}{P}}}}}.
\end{equation*}
Thus, the second judgmental equality we will need is that the diagram
\begin{equation}\label{eq:ealg-eq2}
\begin{tikzcd}[column sep=huge]
\ctxext{P}{{\jcomp{}{e}{P}}{\jcomp{}{e}{{}{e}{P}}}} 
  \ar{r}{\jvcomp{}{\idtm{P}}{\jcomp{}{e}{f}}}
  \ar{d}[swap]{
    \jvcomp{}{f}{\idtm{\jcomp{}{f}{{}{e}{P}}}}
%    \tmext
%      {\ctxwk{\jcomp{}{\bar{e}}{{}{e}{P}}}{f}}
%      {\idtm{\jcomp{}{\bar{e}}{{}{e}{P}}}}
    }
& \ctxext{P}{\jcomp{}{e}{P}} \ar{d}{f}\\
\ctxext{P}{\jcomp{}{e}{P}} \ar{r}[swap]{f} & P
\end{tikzcd}
\end{equation}
commutes judgmentally. Now we can confidently formulate the definition of
extension algebras.

\begin{defn}
An \emph{extension algebra in context $\Gamma$} is a quadruple $(A,P,e,f)$
consisting of a family $A$ over context $\Gamma$, a family $P$ over the context
$\ctxext{\Gamma}{A}$, a morphism $e$ from $\ctxext{A}{P}$ to $A$ in context
$\Gamma$ and a morphism $f$ from $\ctxext{P}{\jcomp{}{e}{P}}$ to $P$ in context
$\ctxext{\Gamma}{A}$, satisfying the judgmental equalities of
\autoref{eq:ealg-eq1,eq:ealg-eq2}.
\end{defn}

We also give a bit of intuition to the requirement of \autoref{eq:ealg-eq2} by
means of the following lemma.

\begin{lem}
Let $(A,P,e,f)$ be an extension algebra in context $\Gamma$ and let
$y_0:P$, $y_1:\subst{y_0}{\jcomp{}{e}{P}}$ and 
$y_2:\subst{y_1}{{y_0}{\jcomp{}{e}{\jcomp{}{e}{P}}}}$. Then we have the
judgmental equality
\begin{equation*}
\subst{{y_2}{{y_1}{\jcomp{}{e}{f}}}}{{y_0}{f}}
  \jdeq
  \subst{y_2}{{{y_1}{{y_0}{f}}}{f}}
\end{equation*}
\end{lem}

\subsection{Extension-empty algebras}
\begin{defn}
Let $P$ be a family over the extended context $\ctxext{\Gamma}{A}$, let
$\jterm{\Gamma}{A}{x}$ and $\jterm{{\Gamma}{A}}{P}{f}$ be terms. Then the
quadruple $(A,P,x,f)$ is said to be an \emph{empty algebra in context $\Gamma$}
if the following judgmental equalities hold:
\begin{align*}
\jfameq*{\Gamma}{\subst{x}{P}}{A}\\
\jtermeq*{\Gamma}{A}{\subst{x}{f}}{x}.
\end{align*}
\end{defn}
